Follow-up Samples

### 2017-04-21 submission/manuscript.Rmd

# The Fecal Microbiome Before and After Treatment for Colorectal Adenoma or Carcinoma
```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

@@

# initial microbiome changes
change_summary <- read.csv('../data/process/tables/change_theta_fit_summary.csv', header = T, row.names = 1)
alpha_summary <- read.csv('../data/process/tables/alpha_table_summary.csv', header = T, row.names = 1)
beta_summary <- read.csv('../data/process/tables/beta_diver_summary.csv', header = T, row.names = 1)
time_pvalues <- read.csv("../data/process/tables/time_pvalues.csv", header = T, row.names = 1)
time_summary <- read.csv("../data/process/tables/time_summary_data.csv", header = T, row.names = 1)
all_OTU_pair_wilcox <- read.csv("../data/process/tables/OTU_paired_wilcoxson_test.csv", header = T)

@@

```
\vspace{25mm}

\begin{center}
Running Title: Human Microbiome before and after Colorectal Cancer

@@

$2$ Department of Internal Medicine, University of Michigan, Ann Arbor, M.

$3$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, P.

@@

**Background.** Colorectal cancer (CRC) is a worldwide health problem. Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to the microbiome after treatment for an adenoma or carcinoma. This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with with disease to those associated with a normal colon. We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), and carcinoma (N=`r change_summary["thetayc", "crc_n"]`).

@@

**Conclusions.** Further exploration of the relationship between diagnosis, treatment, and the impact on the microbiome will yield improvements in disease management.

@@

***Treatment alters the bacterial community structure of patients diagnosed with colonic lesions.***
Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiome of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples (Table 1). The structure of the microbial communities of the pre- and post-treatment samples differed [Figure 1A]. We found that the communities obtained pre- and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001). There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P>0.05). Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1BCD]. We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value`r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value`r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value`r beta_summary["Carcinoma", "PERMANOVA"]`). Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value>0.05) [Table S1].

@@

This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that the microbiome contains biomarkers to diagnosis colonic lesions [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009; @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013]. Our study focused on comparing the microbiome of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment. For all three groups of patients we observed changes in their microbiota. The microbiota of patients diagnosed with carcinoma changed significantly more than the other groups. This change resulted in communities that more closely resembled those of patients with normal colons.

For the carcinoma group there was a larger change between their initial and follow up samples then for either adenoma or advanced adenoma.  This change was directionaly the same for each individual with carcinoma.  This observed change has  important implications for future research into CRC recurrence.  Knowing that the community changes towards normal after treatment means that it is reasonable to hypothesize that the microbiome could be a good biomarker for stratifying individuals between those most and least at risk for CRC recurrence.  Interestingly, our `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability fits nicely within the range quoted for CRC recurrence (20-30%) [@hellinger_reoperation_2006; @ryuk_predictive_2014].

A potential limitation, within our study, is that we do not know whether individuals who had an increased positive probability for carcinoma on follow up eventually had a subsequent CRC recurrence.  This information would help to strengthen the case for the ability of our model to be used as a disease stratification tool for CRC recurrence. This would be an exciting area of future research that would be able to link treatment changes within the microbiome and CRC recurrence together.

The microbiome changes associated with carcinoma were found to be towards those with normal colons.  This would suggest that treatment for CRC is not only successful for removing the carcinoma but also at removing bacteria associated with them.  This is important due to the numerous studies that have pointed towards a link between the community and tumorgenesis [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014].  The importance of the community to this process is further highlighted in our data [Figure 2, 4, & S3].  Interestingly, those who had an increase in their positive probability of carcinoma had no change in their median value for oral pathogens found within the carcinoma model but had a decrease in non-oral pathogens [Figure S4].  Although this finding was not statistically significant, this trend would further highlight the importance of community to tumorgenesis.  Future studies with larger power may be able to confirm these reported observations.

Although we were not able to identify specific OTUs due to the heterogeneity of the population.  This is a common problem that occurs in other fields of microbiome study [@sze_looking_2016].  This made it very difficult to identify individual OTUs associated with treatment and recovery towards a normal colon.  To overcome this large individual heterogeneity we used RF models that made it easier to identify consortia of microbiota that were associated with treatment and recovery.  A potential future direction would be to study whether specific gene signatures associated with these consortia of bacteria have a role to play in the recurrence of CRC.

Despite some of these stated shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable change in the bacterial community after adenoma, advanced adenoma, or carcinoma treatment. Further, individuals with carcinoma had a larger change than adenoma or advanced adenoma indivdiuals and this change in community was towards those with normal colons. Our data provides additional evidence on the importance of the community  on recovery of the microbiome after CRC treatment and opens interesting avenues of research into how these changes may affect recurrence.

@@

***Study Design and Patient Sampling:***
Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016]. Briefly, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis. Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after. The bacterial community has been shown to normalize back to a pre-colonoscopy community within this time period [@obrien_impact_2013]. Our training cohort consisted of a total of `r length(rownames(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)))))` individuals [Table 1]. Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 2]. This study was approved by the University of Michigan Institutional Review Board. All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***16S rRNA Gene Sequencing:***
Sequencing was completed as described by Kozich, et al. [@kozich_development_2013]. DNA extraction used the 96-well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf). The V4 variable region was amplified and the resulting product was split between three sequencing runs with normal, adenoma, and carcinoma evenly represented on each run. Each group was randomly assigned to avoid biases based on sample collection location. The pre- and post-treatment samples were sequenced on the same run.

***Sequence Processing:***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences and has been previously described [@kozich_development_2013]. The general workflow using mothur was: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screened for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm. The number of sequences for each sample was rarefied to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` to minimize uneven sampling.

***Model Building:***
The Random Forest [@breiman_random_2001] algorithm was used to create the model used to create the three models used. The adenoma model classified normal versus adenoma, advanced adenoma was normal versus advanced adenoma, and carcinoma was normal versus carcinoma. The toal number of individuals in this data set was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals. There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` indivdiuals in the carcinoma model [Table 1]. Each model was then applied to our `r length(rownames(good_metaf))`-person cohort testing prediction of adenoma intial (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])`) versus adenoma follow up (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])`), advanced adenoma initial (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])`) versus advanced adenoma follow up (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])`), carcinoma initial (carinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])`) versus carcinoma follow up (carcinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")[, "followUp"])`).

@@

Assessment of the most important OTUs to the model involved counting the number of times an OTU was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different splits run. This was then followed with filtering of this list to variables that were only present in more than 50% of these 100 runs. The final collated list of variables was then run through the mtry optimization again. Once the ideal mtry was found the entire sample set specific to normal versus adenoma, normal versus advanced adenoma, or normal versus carcinoma was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed. For all three models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`

The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion. The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model.

***Statistical Analysis:***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis. Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`). Comparisons between probabilities as well as overall OTU differences between pre- and post-treatment samples utilized a paired Wilcoxson ranked sum test. Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant. Unless otherwise stated the P-values reported are those that were BH corrected. Model rank importance was determined by obtining the median MDA from the 100 20 repeated 10-fold cross validation and then ranking from largest to smallest MDA.

***Analysis Overview:***
We first tested whether there were any differences between pre- and post-treatment samples in alpha and beta diversity based on adenoma, SRN, or carcinoma. We then tested all OTUs for differences between pre- and post-treatment samples. We next used our specific models for adenoma, SRN, and carcinoma to test classification accuracy, response towards a normal microbiome, and common OTUs used across models. Finally, for the adenoma group differences between those that received surgery or not was tested while for the carcinoma group differences between those receiving chemotherapy and radiation was tested.

***Reproducible Methods:***
A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017. Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.

@@

**Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma Models.*. A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification. B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification. C) Carcinoma ROC curve: The light red shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark red line represents the model using 100% of the data set and what was used for subsequent classification.

**Figure S3: Summary of Important Variables for the Adenoma, Advanced Adenoma, and Carcinoma Models.*. A) MDA of the most important variables in the adenoma model. The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs. B) Summary of Important Variables in the advanced adenoma model. MDA of the most important variables in the SRN model. The dark yellow point represents the mean and the lighter yellow points are the value of each of the 100 different runs. C) MDA of the most important variables in the carcinoma model. The dark red point represents the mean and the lighter red points are the value of each of the 100 different runs.

>>>>

```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

@@

# initial microbiota changes
change_summary <- read.csv('../data/process/tables/change_theta_fit_summary.csv', header = T, row.names = 1)
alpha_summary <- read.csv('../data/process/tables/alpha_table_summary.csv', header = T, row.names = 1)
beta_summary <- read.csv('../data/process/tables/beta_diver_summary.csv', header = T, row.names = 1)
time_pvalues <- read.csv("../data/process/tables/time_pvalues.csv", header = T, row.names = 1)
time_summary <- read.csv("../data/process/tables/time_summary_data.csv", header = T, row.names = 1)
all_OTU_pair_wilcox <- read.csv("../data/process/tables/OTU_paired_wilcoxson_test.csv", header = T)

@@

```
\begin{center}

# The effect of treatment on the microbiota of patients diagnosed with colonic lesions

\vspace{25mm}

@@

$2$ Department of Internal Medicine, University of Michigan, Ann Arbor, MI

$3$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA

@@

**Background.** Colorectal cancer (CRC) is a worldwide health problem. Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to the microbiota after treatment for an adenoma or carcinoma. This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with with disease to those associated with a normal colon. We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), and carcinoma (N=`r change_summary["thetayc", "crc_n"]`).

@@

**Conclusions.** Further exploration of the relationship between diagnosis, treatment, and the impact on the microbiota will yield improvements in disease management.

@@

***Treatment alters the bacterial community structure of patients diagnosed with colonic lesions.***
Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples (Table 1). The structure of the microbial communities of the pre- and post-treatment samples differed [Figure 1A]. We found that the communities obtained pre- and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001). There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P>0.05). Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1BCD]. We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value`r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value`r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value`r beta_summary["Carcinoma", "PERMANOVA"]`). Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value>0.05) [Table S1].

@@

This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that the microbiota contains biomarkers to diagnosis colonic lesions [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009; @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013]. Our study focused on comparing the microbiota of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment. For all three groups of patients we observed changes in their microbiota. The microbiota of patients diagnosed with carcinoma changed significantly more than the other groups. This change resulted in communities that more closely resembled those of patients with a normal colon. For the patients diagnosed with carcinomas there was a larger change between their initial and follow up samples than was observed for patients diagnosed with either adenoma or advanced adenoma. The changes in the community structure of those diagnosed with carcinoma were similar and resulted in communities that more closely resembled those found in patients with normal colons.

Understanding the effect of treatment on the microbiota of those initially diagnosed with carcinomas may have important implications into understanding disease recurrence. The microbiota changes associated with treatment among patients initially diagnosed with carcinomas resulted community structures that resembled patients with normal colons. This would suggest that treatment for carcinoma is not only successful for removing the carcinoma but also at removing bacteria associated with them. Interestingly, 6 of the 26 patients diagnosed with carcinomas had an elevated probability of having carcinomas after treatment. This leads us to hypothesize that these individuals may be at a high risk of recurrence. Interestingly, the `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30%), which would support our hypothesis [@hellinger_reoperation_2006; @ryuk_predictive_2014]. Unfortunately, because of the design of our study, it was not possible to determine whether individuals who had an increased positive probability for carcinoma experienced recurrence. Regardless, it is intriguing to think of using biomarkers from the microbiota to not only predict the presence of lesions, but to also assess the risk of developing them.

It is interesting that those patients diagnosed with adenoma and advanced adenoma did not experience a shift towards a community structure that resembled those with normal colons. This may be due to the fundamental differences between the features of adenomas and advanced adenomas and carcinoma. Specifically, carcinomas may create an inflammatory milieu that would impact the structure of the community and removal of that stimulus would alter the structure of the community. In addition, it is possible that the difference between the microbiota of patients with adenoma and advanced adenoma and those with normal colons is subtle. This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas (Figure S2). Given the patchy distribution of microbiota across patients in the different diagnosis groups, it is possible that we lacked the statistical power to adequately characterize the change in the communities following treatment. A final hypothesis is that the specific type of treatment altered the structure of the microbiome. The treatment to remove adenomas and advanced adenomas was either polyp removal or surgical resection whereas it was surgical resection alone or in combination with chemotherapy or with chemotherapy and radiation. Because chemotherapy and radiation target rapidly growing cells, these treatments would be more likely to cause a turn over of the colonic epithelium driving a more significant change in the structure of the microbiota. Although, we were able to test for an effect across these specific types of treatment, the number of patients in each treatment group was relatively small.

It was not clear why 6 of the 26 patients with carcinoma had a higher probability of having carcinomas after treatment. We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these 6 individuals (3 with Stage II and 3 with Stage III) was similar to the distribution observed among the other 20 patients. We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased or decreased (Wilcoxon Test; P-value=0.56). Finally, it is possible that these patients may not have responded to treatment as well as the other 20 patients diagnosed with carcinoma and so the microbiota may not have been impacted the same way. Again, further studies looking at the role of the microbiota in recurrence are needed to understand the dynamics following treatment.

This study builds upon existing work that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions. The most exciting future direction from the current study is the possibility that markers within the microbiota could be used to evaluate the effect of treatment and predict recurrence for those initially diagnosed with carcinoma. If such an approach is effective, it might be possible to even target the microbiota as part of combined therapies including currently standard approaches. Our data provides additional evidence on the importance of the microbiota in tumorigenesis by addressing the recovery of the microbiota after treatment and opens interesting avenues of research into how these changes may affect recurrence.

@@

***Study Design and Patient Sampling.***
Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016]. Briefly, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis. Samples used to build the models for prediction were collected either prior to a colonoscopy or between one and two weeks after initial colonoscopy. The bacterial community has been shown to normalize back to a pre-colonoscopy community within this time period [@obrien_impact_2013]. Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 2]. Patients were diagnosed by colonoscopic examination and histopathological review of biopsies. Patients were classified as having advanced adenoma if they had an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology. This study was approved by the University of Michigan Institutional Review Board. All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***16S rRNA Gene Sequencing.***
Sequencing was completed as described by Kozich, et al. [@kozich_development_2013]. DNA extraction used the 96-well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf). The V4 variable region was amplified and the resulting product was split between three sequencing runs with normal, adenoma, and carcinoma evenly represented on each run. Each group was randomly assigned to avoid biases based on sample collection location. The pre- and post-treatment samples were sequenced on the same run.

***Sequence Processing.***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences and has been previously described [@kozich_development_2013]. The general workflow using mothur included merging paired-end reads into contigs, filtering for low quality contigs, aligning to the SILVA database [reference], screening for chimeras using UCHIME [reference], classifying with a naive Bayesian classifier using the Ribosomal Database Project (RDP)[reference-Wang], and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm [reference-Westcott&SchlossAEM]. The number of sequences for each sample was rarefied to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` to minimize the impacts of uneven sampling.

***Model Building.***

Our training cohort consisted of a total of `r length(rownames(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)))))` individuals [Table 1].

The Random Forest [@breiman_random_2001] algorithm was used to create the model used to create the three models used. The adenoma model classified normal versus adenoma, advanced adenoma was normal versus advanced adenoma, and carcinoma was normal versus carcinoma. The toal number of individuals in this data set was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals. There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` indivdiuals in the carcinoma model [Table 1]. Each model was then applied to our `r length(rownames(good_metaf))`-person cohort testing prediction of adenoma intial (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])`) versus adenoma follow up (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])`), advanced adenoma initial (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])`) versus advanced adenoma follow up (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])`), carcinoma initial (carinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])`) versus carcinoma follow up (carcinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")[, "followUp"])`).

@@

Assessment of the most important OTUs to the model involved counting the number of times an OTU was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different splits run. This was then followed with filtering of this list to variables that were only present in more than 50% of these 100 runs. The final collated list of variables was then run through the mtry optimization again. Once the ideal mtry was found the entire sample set specific to normal versus adenoma, normal versus advanced adenoma, or normal versus carcinoma was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed. For all three models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`. The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model.

***Statistical Analysis.***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis. Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`). Comparisons between probabilities as well as overall OTU differences between pre- and post-treatment samples utilized a paired Wilcoxon ranked sum test. Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant. Unless otherwise stated the P-values reported are those that were BH corrected. Model rank importance was determined by obtaining the median MDA from the 100, 20 repeated 10-fold cross validation and then ranking from largest to smallest MDA.

***Reproducible Methods.***
A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017. Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.

@@

**Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification. B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification. C) Carcinoma ROC curve: The light red shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark red line represents the model using 100% of the data set and what was used for subsequent classification.

**Figure S3: Summary of Important Variables for the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) MDA of the most important variables in the adenoma model. The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs. B) Summary of Important Variables in the advanced adenoma model. MDA of the most important variables in the SRN model. The dark yellow point represents the mean and the lighter yellow points are the value of each of the 100 different runs. C) MDA of the most important variables in the carcinoma model. The dark red point represents the mean and the lighter red points are the value of each of the 100 different runs.
